---
title: "Intro to Vegan"
subtitle: "Written by Gordon Custer 8/28/18"
output: html_notebook
---

As we saw in the phyloseq tutorial some things are not offered and even if they are they may not be exactly what you want. In this tutorial we examine some of the most commonly utilized functions in the Vegan package. The vegan packages was developed by Jari Oksanen and is used for many common ecological analysis. Some of these analyses, such as ordination, we already covered in the phyloseq tutorial. However, to provide you with other options we will also look at their implementation in Vegan. 

Since we have already gone over the basics of ordination in Phyloseq we will not rehash those details here. Instead we will look at an alternative way to do the same analysis. In this tutorial we will also take advantage of base R's graphics. You will see how different a ggplot graphic is from the base R version of the same plot. 

First lets load our data. We should be working with the same R environment we used created from the Phyloseq tutorial. It would be wise to save this as a different file so we don't mess anything up from our phyloseq analysis. 

Exercise: load in your phyloseq R environment and save it under a new name. Remember to save your scripts and environments as something descriptive. Next, install and load the package 'Vegan' and load it. 

Answer:
```{r}
load("~/Desktop/Micro_Div/Phyloseq_Environment.RData")
library(vegan)
```

All analyses in phyloseq utilize the phyloseq object you created at the beginning of the phyloseq tutorial. However, when performing analysis outside of phyloseq you need to deconstruct the phyloseq object. In order to do so you want to pull out the individual parts and save them as new variables. Here I address how to pull out the metadata and ESV table (called the out_table). Remember back to the "Intro to R" tutorial where we briefly mention how to examine what type of data we have and how to change between data structures? Well this is a perfect example of that conversion in real life analysis. 
```{r}
#Remember the "switch"
#ps<-ps_Rhizo_transformed
ps<-ps_Bulk_transformed
ps_table <- as.data.frame(otu_table(ps))
sd <- as.data.frame(as.matrix(as.data.frame(sample_data(ps_Bulk_transformed))))
```

In the vegan package we use the function metaMDS. Look at the help page for the metaMDS. You will notice phyloseq also offers a function called metaMDS. This is likely because phyloseq contains many vegan functions. The nice thing about using metaMDS in phyloseq is that you have the ability to alter many of the arguments that you do not have access to in phyloseq. Here we ordinate our OTU table to visualize community level differences. 

Within the help page titled vegan::metaMDS look at the available options. What does the "k=" argument do? What about the "previous.best=T" argument? 
```{r}
ord<-metaMDS(ps_table, distance = "bray", k=3, trymax = 500, previous.best = T)
```
In the phyloseq tutorial, you were shown how to plot the ordination with the function plot_ordination(). In this chunk the most basic way to make an ordination plot is used. The function plot() is included in base R. You can use it to plot many different types of data. The nice thing about this plot() function is that it will interpret what type of graph is best suited for your data. 

Look at the plot made by plot(ord). Then look at the data made by plot(sd$pH). Notice the same function was used but the graphs are totally different based upon the structure of the input. This is one of the really cool features of R. 
```{r}
plot(ord)
plot(sd$pH)
```

In this section we save the ordination plot as an object named fig1. Look at the help page for the ordiplot() function. What does the "type=" argument mean? We have points as the type in this function. Re-run the function changing the type argument to one of the other options. Can you think of the advantages of having a figure with points instead of text? What about the other way around?

Look at the ordiplot function and examine what other 'type' you could use. Try plotting with some of he others. 
```{r}
fig1<-ordiplot(ord, type = "points")
fig1
```

In order to beef up our plot we can try a couple different things. Much like the phyloseq object has other objects contained within it, fig1 has contains objects as well. To look at what fig1 contains use the "$". Remember the "$" was used to pull out specific data from an object. Press tab when your cursor is after the "$" and R will provide the available options. 
```{r}
fig1$
```

The sites object gives the location of the sites. What we are doing here is pulling this data out and saving at is a new variable. We are doing this so we can plot the locations and have slightly more control over the final plot. 
```{r}
NMDS1<-fig1$sites[,1]
NMDS1
NMDS2<-fig1$sites[,2]
NMDS2
```

A useful feature of the Vegan package is the ability to plot environmental vectors and factors over an ordination. The projection of the vector in relation to the points in the ordination shows the correlation of those points with that environmental vector. These environmental variables can include thing such as pH, Moisture, C:N or could even be enzyme activities and they make inference and interpretation much easier. Below is some code which produces a graph with environmental variables laid over. One important thing to note about using R markdown is that base R graphics must be ran together in the same chunk. If you were to split this up and run the plot, the try to add the title you will get an error. 
```{r}
plot(NMDS1, NMDS2,cex=0.8, xlab='NMDS1', ylab='NMDS2')
with(sd,ordiellipse(ord,Infestation_Stage,kind="se",conf=0.95))

#to plot only variable with a certain p-value (p.max)
ord.fit <- envfit(ord~Infestation_Stage + pH + Gravimetric_Water, data = sd,  perm = 1000, na.rm = TRUE)
plot(ord.fit, p.max=0.05, col="black", cex=0.8)
title(main = 'Bulk Soil Bacterial Community', cex.main=2)
```

Next up is Adonis testing. Adonis testing is a permutation based MANOVA. This means we can test for significance of treatments or uncontrolled covariates. The help page has a very good explanation of what the adonis() function does. Check this out.

```{r}
adonis_table<-data.matrix(ps_table)
adonis_dist<-parDist(adonis_table, method = "bray")
adonis(ps_table)

ord<-metaMDS(adonis_table, distance = "bray", k=3, trymax = 1000, previous.best = T)
#psNMDS <- ordinate(ps_adonis, "NMDS", psdist)

adonis(adonis_dist ~  Infestation_Stage , data = sd)
adonis(adonis_dist ~ Infestation_Stage + pH + Gravimetric_Water, data = sd)

```


Lets assume you are interested in how a soil microbial community changes with respect to the edaphic conditions present at a site. Your hypothesis might be as edaphic conditions become increasingly different so will the microbes that are present. But how would you test this? Any ideas?

A mantel test produces a mantel test statistics which is the correlation between two dissimilarity matrices. Be careful to make sure you samples are in the same order for both matrices! If they are not you will not be comparing the correct samples to each other. 

To run the mantel test we want to keep only samples in both data frames. We do this because we cant compare a sample if it doesn't have a match. First we want to pull out the sample names of the two data frames and store them in a vector
```{r}
ps_table<-na.omit(ps_table)
sd<-na.omit(sd)
PS_samples<-rownames(ps_table)
Metadata_Samples<-rownames(sd)
```

The first part here we match the two vectors of site names we made above. See how the %in% command is used. This is extremely useful when trying to match values in two vectors or data frames.
```{r}
PS_samples<-PS_samples[PS_samples %in% Metadata_Samples]
Metadata_Samples<-Metadata_Samples[Metadata_Samples %in% PS_samples]
```

Next once we have matched the values of our two vectors, we can then filter our data frames to include only samples which are present in both. Do you notice how we accomplished this? First we pulled out all the sample names from both data frames. Next, we matched the vectors using the %in% command to remove samples which are not present in both. Finally, we then did the same for the entire data frames. The end product is two data frames with the same samples. 
```{r}
PS_matched<-ps_table[rownames(ps_table) %in% Metadata_Samples,]
MD_Matched<-sd[rownames(sd) %in% PS_samples,]
```

Now that you think you have filtered the samples it is good practice to check to make sure it matches. True indicates a match, while false a non-match. 
```{r}
rownames(PS_matched) == rownames(MD_Matched)

```
Exercise: Can you think of a way to check whether or not everything matches in a huge data frame using the unique function?  

Answer:
```{r}
unique(rownames(PS_matched) == rownames(MD_Matched))
```
Now that we have subset our data frame to only include samples which match we can calculate the distance between samples. The function parallelDist() allows for parallel processing when computing distance matrices. This is extremely important because large data sets can take hours to process if processes on a single core. However, this function is picky about the input data type. Lets look at the help page and see what it has to say. What type of data (structure) is required by parallelDist()?

Something else to keep in mind is that when computing a distance matrix we will want only numeric values in our data frame. Here we convert the object to a data frame, create row name based on the sample names, remove the sample names, and then convert the data frame to the correct type, a matrix. 
```{r}
PS_matched<-as.matrix(PS_matched)
MD_Matched<-as.matrix(MD_Matched)
```

Once you have your data in the correct format and you know that your samples are in the correct order you can proceed! Using the parallelDist() function we calculate a distance matrix. This is the same operation we performed in the phyloseq tutorial with the dist() function. Take a look at the object once it has finished running. What do the numbers mean? This might be a question for Google!
```{r}
bray_PS_Matched<-parallelDist(PS_matched, method = "bray")
```
The parallelDist() function only accepts numeric inputs. However, our metadata contains both numeric and categorical data. We will have to subset our metadata to only include the numeric variables. Actually, we will want to subset our data to only include abiotic soil conditions. You can accomplish this by calling only certain columns like we did in the earlier tutorials. 

Exercise: Create a data frame that contains only the edaphic/abiotic soil characteristics. With this data frame run the parallelDist function. This is a tricky exercise and may require some serious Goggling!

Answer:
```{r}
edaphic_measurements<-as.data.frame(MD_Matched[,c(2:4,16)])
edaphic_measurements<-lapply(edaphic_measurements, as.character)
edaphic_measurements<-data.frame(lapply(edaphic_measurements, as.numeric))
rownames(edaphic_measurements)<-rownames(MD_Matched)
edaphic_measurements<-as.matrix(edaphic_measurements)

bray_edaphic_Matched<-parallelDist(edaphic_measurements, method = "bray")
```


Now for the statistics. A mantel test works by permuting one of the matrices and comparing the correlation to the original. The idea is that if the null hypothesis is true and there is no relation between the two matrices then shuffling (permuting) the one matrix will produce correlations larger and smaller than the correlation between the two original distance matrices. However, if a the two distance matrices are in fact correlated then permuting one will produce a smaller correlation many times and a larger correlation fewer times. If the "true" observed statistics is larger than 95% (0.05) of the permuted statistics then the test we can reject the null hypothesis. 

Let's put this another way. If the statistic is very large and we accept the alternative hypothesis then we can say the two distance matrices are related and a change in one will result in a change in the other. An example of which would be if we saw pH change and a resulting change in bacteria community (which we should all know is the case!). 

Run this mantel test and examine the output. What conclusions can you make? 
```{r}
mantel(bray_edaphic_Matched, bray_PS_Matched, method = "pearson")
```

We can also plot the distance matrices but this becomes a bit more difficult. In order to do so, we need to extract the data from the distance matrices reformat it and then plot. The best way to plot this type of data will be a scatter plot. Lets examine how you might go about this. 

Since we already have out distance matrix lets make a switch to call them data and data2
```{r}
dist_data<-bray_edaphic_Matched
dist_data2<-bray_PS_Matched
```
Next we will work on changing the distance matrices from triangular format like they are normally to a list. This can take some coercing and was found after a long evening of searching the internet. 

First, make your distance matrix into a list of observations and distance between two points then name your columns. The first two columns will be the sites the third will be the bray-Curtis distance between the two sites this can be used to call the columns if you wanted to pull certain variables from your site names. 
```{r}
data_pairs<- data.frame(t(combn(rownames(edaphic_measurements),2)), as.numeric(dist_data))
names(data_pairs) <- c("Site1", "Site2", "distance")
data2_pairs<- data.frame(t(combn(rownames(PS_matched),2)), as.numeric(dist_data2))
names(data2_pairs) <- c("Site1", "Site2", "distance")
```
You will want to make sure the order of samples matches up here. 
```{r}
data_pairs$Site1 == data2_pairs$Site1
data_pairs$Site2 == data2_pairs$Site2
```
They should all match. Do you remember how to use the unique command to check that all values are true? 

Exercise: Using the unique function check to make sure all sites match. 

Next we want to combine the data frames so we have two points for each pair of values. One for the distance between bulk samples and one for the distance between the rhizosphere samples. 
```{r}
combined_data_frame <- data.frame(delta_bulk = data_pairs$distance, delta_rhizo = data2_pairs$distance, Site1 = data2_pairs$Site1, Site2 = data2_pairs$Site2)
```

Now we plot. This is done using ggplot.
```{r}
p<-ggplot(data = combined_data_frame, aes(x=delta_bulk, y=delta_rhizo)) + geom_point()
p <-p + theme(panel.background = element_rect(fill='white', colour='black'))
p <- p + ylab("Change in Bulk Community")
p <- p + xlab("Change in Abiotic Conditions")
p <- p + theme(plot.title = element_text(size = 30 ,hjust = 0.5))
p <- p +theme(axis.title.x=element_text(size=15))
p <- p +theme(axis.title.y=element_text(size=15))
p <- p + guides(fill=guide_legend(title=""))
#p <- p + annotate("Add text here", x = , y = )
p
```
With the last line you could include the Mantel test statistic and p-value for the mantel test you ran above. This would be a great example of a publication ready graphic. 
